---
title: "Reference semantics -- data.table"
output: 
   html_document:
    toc: yes
    toc_depth: 3
    css: ../style.css
params:
  date: !r Sys.Date()    
---

```{r, warning=FALSE, message=FALSE}
library(knitr)
library(data.table)
```

```{r,setup, include=FALSE, eval=TRUE}
options(knitr.table.format = "html", width = 140)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 12, fig.height = 8)
```

<div>Author: Rick Dean</div>
<div>Article date: `r params$date`</div>

<div class="abstract">
  <p class="abstract">Abstract</p>
The following contains notes/R scripts on learning data.table from the R package vignette [Reference semantics](https://cloud.r-project.org/web/packages/data.table/vignettes/datatable-reference-semantics.html)   
</div>  

# Data
## Read in the NYC-flights14 data

<div class="note">Note: As referenced in the vignette, we have downloaded the [flight data](https://raw.githubusercontent.com/Rdatatable/data.table/master/vignettes/flights14.csv) csv file.</div>
```{r}
nyc_flights14_file <- file.path("E:/airline_data/flights14.csv")
if(!base::file.exists(nyc_flights14_file)){
  stop("Source airline file does not exist.")
}
flights <- data.table::fread(nyc_flights14_file)
str(flights)
```

# Reference semantics
## The `:=` operator
<blockquote>With data.table’s `:=` operator, absolutely no copies are made..., irrespective of R version you are using. This is because `:=` operator updates data.table columns in-place (by reference).</blockquote>

The `:=` can be used in two ways in `j`:

a. The LHS `:=` RHS form:
<div><code class = "sourceCode">
DT[, c("colA", "colB", ...) := .(valA, valB, ...)]
</code></div>

b. The functional form:
<div><code class = "sourceCode">
DT[, := (colA = valA, <a># valA is assigned to colA</a><br>
<span>       colB = valB, ...)]</span> <a># valB is assigned to colB</a>
</code></div>

<div class="note">Note: With the functional form you can add comments.</div>

## Add columns by reference
How can we add columns *speed* and *total_delay* of each flight?
```{r}
flights[, `:=`(speed = distance/(air_time/60),  # speed in mph (mi/h)
               total_delay = arr_delay + dep_delay)] # total_delay in minutes
head(flights)
```
<div class="note">Note: Using the functional form here.</div>

## Update some rows of columns
Look at the *hours* available values:
```{r}
flights[, sort(unique(hour))]
```
Replace 24 with 0 using subassign by reference:
```{r}
flights[hour == 24L, hour := 0L][1:3]
```
<div class="note">Note: Using the LHS `:=` RHS form and the update takes place invisibly.</div>

Look at the *hour* variable again to verify:
```{r}
flights[, sort(unique(hour))]
```
## Using a character variable with `:=`
1. Define a simple data.table:
```{r}
dt <- data.table::data.table(mtcars)[, .(cyl, mpg)]
str(dt)
```
2. Add a new column called *mpg_2* by assigning the name to a variable:
```{r}
new_var_name <- "mpg_2"
```

3. Use the variable name with `:=` by just surrounding the variable with parenthesis:
```{r}
dt[, (new_var_name) := mpg * 2]
str(dt)
```


## Delete column by reference
Remove *total_delay* column that we added above:
```{r}
flights[, `:=`(total_delay = NULL)][1:3]   # delete delay column
```
## Using `:=` along with `by`
How can we add a new column which contains for each *orig*, *dest* pair the maximum speed?
```{r}
flights[, `:=`(max_speed = max(speed)), by = .(origin, dest)][1:6]
```
## Using `.SD` to add two columns for each grouping of *month*
```{r}
in_cols <- c("dep_delay", "arr_delay")
out_cols <- c("max_dep_delay", "max_arr_delay")
flights[, c(out_cols) := lapply(.SD, max), by = month, .SDcols = in_cols][1:6]
```
<div class="note">Note: If we had used `flights[, out_cols := lapply(.SD, max)` then we would have been creating a new column named *out_cols*. We allow assignment by reference without quoting column names when there is only one column.  That is why we must wrap *out_cols* with `c( )` to differentiate this case.</div>

Delete the newly added columns *speed*, *max_speed*, *max_dep_delay*, *max_arr_delay* to set up for the next sections:
```{r}
flights[, c("speed", "max_speed", "max_dep_delay", "max_arr_delay") := NULL][1:3]
```
# `:=` and `copy()`
<blockquote>:= modifies the input object by reference. Apart from the features we have discussed already, sometimes we might want to use the update by reference feature for its side effect. And at other times it may not be desirable to modify the original object, in which case we can use copy() function, as we will see in a moment.</blockquote>

## `:=` for its side effect
Write a simple function to do two modifications:
```{r}
foo <- function(DT){
  DT[, speed := distance / (air_time/60)] # adding a new column to the inputted data.table
  DT[, .(max_speed = max(speed)), by = month] # returning a new data.table with columns max_speed and grouping variable month
}
ans <- foo(flights)
```
The function returns a new data.table with columns *max_speed* and grouping variable *month*:
```{r}
ans[1:3]
```
Since the function is referencing the same data.table `flights` in memory, a new column *speed* is added to `flights`:
```{r}
flights[1:3]
```
## The `copy()` function
Sometimes we want to pass a copy of the data.table to the function and not a reference to its memory location.
The `copy()` function makes a **deep** copy of the input data.table.

Redefine the function where we will make a copy of the referenced data.table:
```{r}
flights[, speed := NULL]
foo <- function(DT){
  DT <- data.table::copy(DT)
  DT[, speed := distance / (air_time/60)] # adding a new column to the copied data.table
  DT[, .(max_speed = max(speed)), by = month] # returning a new data.table with columns max_speed and grouping variable month
}
```
Show the returned data.table:
```{r}
ans <- foo(flights)
ans[1:3]
```
Show `flights` which should not have the *speed* column:
```{r}
flights[1:3]
```
An additional case for using `copy()` is when we assign column names to a variable and then `add/update/delete` columns by reference. The variable with the names will also be modified.  To illustrate:
```{r}
DT <- data.table(x = 1L, y = 2L)
DT_names <- names(DT)
DT_names
```
Now add a new column to `DT`:
```{r}
DT[, z := 3L]
DT
```
Show `DT_names`:
```{r}
DT_names
```
Use `copy()` and add a new column to `DT` data.table:
```{r}
DT_names = copy(names(DT))
DT[, w := 4L]
```
Show `DT_names` which does not get updated:
```{r}
DT_names
```



