---
title: "Keys and Fast Binary Search Based Subsets"
output: 
   html_document:
    toc: yes
    toc_depth: 3
    css: ../style.css
params:
  date: !r Sys.Date()    
---

```{r, warning=FALSE, message=FALSE}
library(knitr)
library(tictoc)
library(data.table)
```

```{r,setup, include=FALSE, eval=TRUE}
options(knitr.table.format = "html", width = 140)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 12, fig.height = 8)
```

<div>Author: Rick Dean</div>
<div>Article date: `r params$date`</div>

<div class="abstract">
  <p class="abstract">Abstract</p>
  The following contains notes/R scripts on learning data.table from the R package vignette [Keys and fast binary search based subsets](https://cloud.r-project.org/web/packages/data.table/vignettes/datatable-keys-fast-subset.html). 
</div>  

# Data
## Read in the NYC-flights14 data

<div class="note">Note: As referenced in the vignette, we have downloaded the [flight data](https://raw.githubusercontent.com/Rdatatable/data.table/master/vignettes/flights14.csv) csv file.</div>
```{r}
nyc_flights14_file <- file.path("E:/airline_data/flights14.csv")
if(!base::file.exists(nyc_flights14_file)){
  stop("Source airline file does not exist.")
}
flights <- data.table::fread(nyc_flights14_file)
str(flights)
```

# Keys
## What is a `key`?
All data.frames have a row names attribute. Consider the data.frame `DF` below:
```{r}
set.seed(1L)
DF = data.frame(ID1 = sample(letters[1:2], 10, TRUE),
                ID2 = sample(1:3, 10, TRUE),
                val = sample(10),
                stringsAsFactors = FALSE,
                row.names = sample(LETTERS[1:10]))
DF
```
Show the row names of `DF`:
```{r}
rownames(DF)
```
Subset `DF` by referencing a row:
```{r}
DF["C",]
```
Limitations of row names:

1. Each row is limited to **exactly** one row name.
2. Row names should be unique.

Now convert `DF` to a data.table:
```{r}
DT <- as.data.table(DF)
DT
```
Show the row names of data.table `DT`:
```{r}
rownames(DT)
```
Note the following:

1. The row names have been reset from letters to numeric characters.
2. data.tables never uses row names.
3. To preserve the original row names use `keep.rownames = TRUE` in `as.data.table()`. This creates a new column called *rn* with the original row names.

<blockquote>
Keys and their properties

1. We can set keys on multiple columns and the column can be of different types – integer, numeric, character, factor, integer64 etc. list and complex types are not supported yet.

2. Uniqueness is not enforced, i.e., duplicate key values are allowed. Since rows are sorted by key, any duplicates in the key columns will appear consecutively.

3. Setting a key does two things:

  a. physically reorders the rows of the data.table by the column(s) provided by reference, always in increasing order.

  b. marks those columns as key columns by setting an attribute called **sorted** to the data.table.

Since the rows are reordered, a data.table can have at most one key because it can not be sorted in more than one way.
</blockquote>

## Set keys on a data.table
How can we set the column *origin* as key in the data.table `flights`?
```{r}
data.table::setkey(flights, origin)
flights[1:3]
```
<div class="note">Note: The data.table is now reordered via the *origin* column. Also when creating data.tables with `data.table()` you can set the *key* argument which takes a character vector of column names.</div>

## Use keys on a data.table
Use the key column *origin* to subset all rows where the origin airport is "JFK":
```{r}
ans <- flights[.("JFK")]
str(ans)
```
Use the key column *origin* to subset all rows where the origin airport is "JFK" or "LGA":
```{r}
ans <- flights[.(c("JFK", "LGA"))]
str(ans)
```
Check the result values for *origin*:
```{r}
ans[, sort(unique(origin))]
```
## Get the names of the columns keyed
```{r}
data.table::key(ans)
```
## Keys and multiple columns
<blockquote>To refresh, keys are like supercharged row names. We can set key on multiple columns and they can be of multiple types.</blockquote>

How can I set keys on both *origin* and *dest* columns?
```{r}
setkey(flights, origin, dest)
key(flights)
```
Set keys by a character vector:
```{r}
setkeyv(flights, c("origin", "dest"))
```

Subset all rows where the *origin* key is "JFK" and the *dest* key is "MIA":
```{r}
ans <- flights[.("JFK", "MIA")]
str(ans)
```
Check result *origin* values:
```{r}
ans[, sort(unique(origin))]
```
Check result *dest* values:
```{r}
ans[, sort(unique(dest))]
```
Subset all rows where just the first key column *origin* matches "JFK":
```{r}
ans <- flights[.("JFK")]
str(ans)
```
Check the result values for *Origin*:
```{r}
ans[, sort(unique(origin))]
```

Subset all rows where just the second key column *dest* matches "MIA":
```{r}
ans <- flights[.(unique(origin), "MIA")]
str(ans)
```

Check the values for *dest*:
```{r}
ans[, sort(unique(dest))]
```
<div class="note">Note: "MIA" is automatically recycled to fit the unique values of *origin*.</div>

## Subset with keys and select columns with `j` 
Select columns *origin*, *dest*, and *arr_delay* with `j` and subset with keys *origin* = "LGA" and *dest* = "TPA":
Check the keys:
```{r}
key(flights)
```
```{r}
ans <- flights[.("LGA", "TPA"), .(arr_delay, origin, dest)]
str(ans)
```

## Chaining keys with `j`
Use chaining in the above *origin* = "LGA" and *dest* = "TPA" subsetting with ordering by *arr_delay*:
```{r}
ans <- flights[.("LGA", "TPA"), .(arr_delay, origin, dest)][order(-arr_delay)]
str(ans)
```

## Subset with keys and perform a computation in `j`
```{r}
ans <- flights[.("LGA", "TPA"), max(arr_delay)]
ans
```
Note that this value agrees with the first value for *arr_delay* in the previous chunk code above.

## Subset with keys and update column values with `:=` in `j`
Get the unique values for *hour*:
```{r}
flights[, sort(unique(hour))]
```
Replace all values of 24 with 0 using *hour* as a key.
Set the key:
```{r}
setkey(flights, hour)
key(flights)
```
Subset with the key values equal to 24 and set their value to 0 in `j`:
```{r}
flights[.(24), hour := 0L]
```
Check the values for *hour*:
```{r}
flights[, sort(unique(hour))]
```
Check the key of `flights`:
```{r}
key(flights)
```
<div class="note">Note: With the change in *hour* it is no longer sorted and the data.table's key is set to NULL automatically.</div>

## Subset with keys and using `by`
Set the key to *origin* and *dest*:
```{r}
setkey(flights, origin, dest)
key(flights)
```
Get the maximum departure delay for each *month* cooresponding to *origin* = "JFK". Order the results by *month*.
```{r}
ans <- flights["JFK", .(max_dep_delay = max(dep_delay)), keyby = month]
ans[1:6]
```
Check the key for the result:
```{r}
key(ans)
```
<div class="note">Note: `keyby` automatically reset the key to *month*</div>

## Subset with keys and use `mult`
With `mult` values we can select "first", "last", or "all" within a key subset.
Subset only the first matching row from all rows where *origin* matches "JFK" and *dest* matches "MIA":
```{r}
setkey(flights, origin, dest)
ans <- flights[.("JFK", "MIA"), mult = "first"]
ans
```

Subset only the last matching row of all the rows where *origin* matches "LGA", "JFK", "EWR" and *dest* matches "XNA":

```{r}
ans <- flights[.(c("LGA", "JFK", "EWR"), "XNA"), mult = "last"]
ans
```
## Subset with keys and use `nomatch`
We can choose if queries that do not match should return NA or be skipped altogether.
From the previous example:
```{r}
ans <- flights[.(c("LGA", "JFK", "EWR"), "XNA"), mult = "last", nomatch = NULL]
ans
```

# Binary search vs vector scans
Create a sample data.table with 20 million rows and three columns and key it by columns *x* and *y*:
```{r}
set.seed(2L)
N = 2e7L
DT <- data.table(x = sample(letters, N, T),
                 y = sample(1000L, N, T),
                 val = runif(N))
print(object.size(DT), units = "Mb")
```
Time the vector scan approach:
```{r}
tictoc::tic("vector scan")
ans1 <- DT[x == "g" & y == 877L]
tictoc::toc()
```
The vector scan result:
```{r}
str(ans1)
```



Time the binary search approach:
```{r}
setkeyv(DT, c("x","y"))
key(DT)
tictoc::tic("binary search")
ans2 <- DT[.("g", 877L)]
tictoc::toc()
```
The binary search result:
```{r}
str(ans2)
```
Are the *val* results identical?
```{r}
identical(ans1$val, ans2$val)
```
The speed up is ~ `r 0.5/0.01` times
