---
title: "Introduction to data.table"
output: 
   html_document:
    toc: yes
    toc_depth: 3
    css: ../style.css
params:
  date: !r Sys.Date()    
---

```{r, warning=FALSE, message=FALSE}
library(knitr)
library(data.table)
```

```{r,setup, include=FALSE, eval=TRUE}
options(knitr.table.format = "html", width = 140)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 12, fig.height = 8)
```

<div>Author: Rick Dean</div>
<div>Article date: `r params$date`</div>

<div class="abstract">
  <p class="abstract">Abstract</p>
The following contains notes/R scripts on learning data.table from the R package vignette [Introduction to data.table](https://cloud.r-project.org/web/packages/data.table/vignettes/datatable-intro.html) 
</div>  

# Data
## Read in the NYC-flights14 data

<div class="note">Note: As referenced in the vignette, we have downloaded the [flight data](https://raw.githubusercontent.com/Rdatatable/data.table/master/vignettes/flights14.csv) csv file.</div>
```{r}
nyc_flights14_file <- file.path("F:/airline_data/flights14.csv")
if(!base::file.exists(nyc_flights14_file)){
  stop("Source airline file does not exist.")
}
flights <- data.table::fread(nyc_flights14_file)
str(flights)
```

# Basics

## Create a data.table from scratch
```{r}
DT = data.table(
  ID = c("b","b","b","a","a","c"),
  a = 1:6,
  b = 7:12,
  c = 13:18
)
DT
class(DT)
```

<div class="note">Note: data.table syntas:</div>
DT[i, j, by]  Take DT, subset/reorder rows using i, update/perform expressions/select columns using j, group with `by =`.

## Get a subset of rows in `i`
```{r}
ans <- flights[origin == "JFK" & month == 6L]
ans[1:6]
```
## Get the first two rows
```{r}
flights[1:2]
```

## Using a logical vector/column to select rows
```{r}
DT = data.table(
  ID = c("b","b","b","a","a","c"),
  a = 1:6,
  b = 7:12,
  c = 13:18
)
DT
```
Define a logical vector to select rows from the data.table *DT*:
```{r}
take = c(F, T, F, T, F, T)
```
Use the logical vector to subset *DT*:
```{r}
DT_sub <- DT[take]
DT_sub
```
Define a data.table with a logical column:
```{r}
DT = data.table(
  ID = c("b","b","b","a","a","c"),
  a = 1:6,
  b = 7:12,
  c = 13:18,
  take = c(F, T, F, T, F, T)
)
DT
```
Use the logical column to select rows with a negate sign:
```{r}
DT[!take]
```

## Sort *origin* in ascending order, then *dest* in descending order
```{r}
ans <- flights[order(origin, -dest)]
ans[1:6]
```

## Select column(s) in `j` and return it as a vector
```{r}
ans <- flights[, arr_delay]
ans[1:6]
```
The class of the result is an integer vector:
```{r}
class(ans)
```

## Select a column in `j` and return a data.table
```{r}
ans <- flights[,.(arr_delay)]
ans[1:6]
```
The class of the result is a `data.table`:
```{r}
class(ans)
```

## Select multiple columns in `j` and return a data.table
```{r}
ans <- flights[,.(arr_delay, dep_delay)]
ans[1:6]
```
## Select columns and rename them in `j`
```{r}
ans <- flights[, .(delay_arr = arr_delay, delay_dep = dep_delay)]
ans[1:6]
```
## Compute an expression in `j`
How many trips have had total delay < 0?
```{r}
ans <- flights[, sum((arr_delay + dep_delay) < 0)]
ans
```
<div class="note">Note: Columns are variables and we should be able to **compute** by calling functions on those variables.</div>


## Compute an exprssion in `j` with `lapply()` (EX 1)
```{r}
dt <- data.table(x = 1:2, y = 2:3)

sum_fun <- function(xx = NULL){
  if(3 %in% xx){
    return(xx + 2)
  }else {
    return(xx + 4)
  }
}

mod_cols <- c("x","y")
dt[, (mod_cols) := lapply(dt,sum_fun)]
dt
```

## Compute an exprssion in `j` with `lapply()` (EX 2)
```{r}
dt <- data.table(x = c(1,1), y = 2:3)

length_fun <- function(xx = NULL){
  if(length(unique(xx)) != 1){
    return(xx)
  }else{
    return(NULL)
  }
}

mod_cols <- c("x","y") 
dt[, (mod_cols) := lapply(dt, length_fun)]
dt
```


## Subset in `i` and expressions in `j`
Calculate the average arrival and departure delay for all the flights with "JFK" as the *origin* airport in the month of June.
```{r}
ans <- flights[origin == "JFK" & month == 6L, .(m_arr = mean(arr_delay), m_delay = mean(dep_delay))]
ans
```

## Finding the number of rows using `.N`
How many trips have been made in 2014 from "JFK" airport in the month of June.
Using the length of a column:
```{r}
ans <- flights[origin == "JFK" & month == 6, length(dest)]
ans
```
Using the special `.N` symbol:
```{r}
ans <- flights[origin == "JFK" & month == 6L, .N]
ans
```
<div class="note">Note: 

1. In the above example `.N` returns an integer vector.  If it were `.(.N)` a data.table would be returned.

2. `.N` returns the number of rows in the subset. If `by =` is present then the symbol returns the number of observations *in the current group*. 

3. We could have used `nrow(flights[origin == "JFK" & month == 6L])` but it would have to subset the entire data.table first corresponding to the row indices in `i` and then return the rows using `nrow()`, which is unnecessary and inefficient. </div>

## Selecting columns using explicit column string names
```{r}
ans <- flights[,c("arr_delay", "dep_delay")]
head(ans)
```
## Using column names stored in a variable with the `..` prefix
```{r}
select_cols <- c("arr_delay", "dep_delay")
ans <- flights[,..select_cols]
ans[1:6]
```
## Deselect columns using column names
```{r}
ans <- flights[, !c("arr_delay", "dep_delay")]
ans[1:6]
```
<div class="note">Note: You can also use `-`.</div>
## Select columns by using a range of column names
```{r}
ans <- flights[, day:year]
ans[1:6]
```

# Aggregations
Using `i` and `j` together with `by` to perform operations `by group`.

## Grouping using `by`
How can we get the number of trips corresponding to each origin airport?
```{r}
ans <- flights[, .(.N), by = .(origin)]
ans
```
<div class="note">Note: As mentioned above `.N` gives the number of observations *in the current group* and an automatic column name of "N" is assigned.</div>

`by` can also recognizes the column name string:
```{r}
ans <- flights[, .(.N), by = "origin"]
ans
```
How can we calculate the number of trips for each *origin* airport for carrier code "AA"?
```{r}
ans <- flights[carrier == "AA", .(.N), by = "origin"]
ans
```
## Grouping with reference to a variable name
### Using `get()`
```{r}
col <- c("origin")
ans <- flights[, .(.N), by = get(col)]
ans
```

### Using `mget()`
```{r}
cols <- c("origin","dest")
ans <- flights[, .(.N), by = mget(cols)]
ans
```

## Grouping by a pair of columns
How can we get the total number of trips for each *origin* and *dest* combination for carrier code "AA"?
```{r}
ans <- flights[carrier == "AA", .(.N), by = .(origin, dest)]
ans[1:6]
```
## Grouping and performing a statistic
How can we get the average arrival and departure delay for each *origin* and *dest* combination for each month for carrier code "AA"?
```{r}
ans <- flights[carrier == "AA", .(carrier, .N, avg_arr_delay = mean(arr_delay), avg_dep_delay = mean(dep_delay)), by = .(origin, dest, month)]
ans[1:6]
```
<div class="note">Note: The order of the observations in each group is preserved in the output.</div>

## Using `by:keyby` to sort the within the grouping
If you want to order the observations in each group in ascending order then use `keyby = .()`.
```{r}
ans <- flights[carrier == "AA", .(avg_arr_delay = mean(arr_delay), avg_dep_delay = mean(dep_delay)), keyby =.(origin, dest, month)]
ans[1:6]
```
## Chaining
How can we get the number of by *origin* and *dest* and then order these totals?  Answer--use chaining.
```{r}
ans <- flights[carrier == "AA", .(.N), by = .(origin, dest)][order(origin, -dest)]
ans[1:6]
```
An alternative chaining method using `magrittr::` pipes:
```{r}
library(magrittr)
ans <- flights %>% 
  .[carrier == "AA"] %>% 
  .[, .(.N), by = .(origin, dest)] %>% 
  .[order(origin, -dest)]
ans[1:6]
```
## Expressions in `by`
Can `by` accept expressions as well or does it just take columns?
How many flights started late but arrived early (or on time), started and arrived late etc...?
```{r}
ans <- flights[, .(.N), by = .(dep_delay > 0, arr_delay > 0)]
ans
```
## Using the special symbol `.SD` for computations within each group
`.SD` stands for **S**ubset of **D**ata.  It by itself is a `data.table` that holds the data for *the current group* defined using `by`.
### Taking the above `data.table` *DT*, group by "ID" and print each subgroup.
1. Show DT:
```{r}
DT
```
2. Print the subgroups:
```{r}
DT[, print(.SD), by = ID]
```

<div class="note">Note: `.SD` contains all the columns except the grouping columns by default and all the rows for each individual group.</div>

### Using the base lapply to compute the means within each group:
```{r}
DT[, lapply(.SD, mean), by = ID]
```
If we group flights by *origin*, *dest*, *month* for *carrier* = "AA", find the group means for just *dep_delay* and *arr_delay*.
```{r}
ans <- flights[carrier == "AA",
               lapply(.SD, mean),
               by = .(origin, dest, month),
               .SDcols = c("dep_delay", "arr_delay")]
ans[1:6]
```
### Using `.SD` return the first two rows for each *month*.
```{r}
ans <- flights[, .SD[1:2], by = month]
ans[1:8]
```
### Using `.SD` return the first row for each *month*.
```{r}
ans <- flights[, .SD[1L], by = month]
ans[1:8]
```
### Using `.SD` return the last row for each *month*.
```{r}
ans <- flights[, .SD[.N], by = month]
ans[1:8]
```

### Using .SD return a random row of each *month*.
```{r}
ans <- flights[, .SD[sample(.N, 1L)], by = month]
ans[1:8]
```
<div class="note">Note: in `sample(.N, 1L)` sampling takes place from 1:.N; 1L for the number of items to choose. </div>

### Using .SD to select the desired row **dynamically**
Get the maximum *dep_delay* from each carrier:
```{r}
ans <- flights[, .SD[which.max(dep_delay)], by = carrier]
ans[1:8]
```
### Use .SD to sort each carrier group by descending *dep_delay* and take the top 3
```{r}
ans <- flights[, setorder(.SD, -dep_delay)[1:3], by = carrier]
ans[1:12]
```

