---
title: "A dummies guide to the fabulous data table package"
output: 
   html_document:
    toc: yes
    toc_depth: 3
    css: ../style.css
params:
  date: !r Sys.Date()      
---

```{r,setup, include=FALSE, eval=TRUE}
options(knitr.table.format = "html")
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 12, fig.height = 8)
```

<div>Author: Rick Dean</div>
<div>Article date: `r params$date`</div>

<div class="abstract">
  <p class="abstract">Abstract</p>
The following notes/R scripts are inspired on an article by John MacKintosh, [data.table by a dummy](https://johnmackintosh.net/blog/2019-06-30-datatable-by-a-dummy/)
</div>

```{r, warning=FALSE, message=FALSE}
library(knitr)
library(data.table)
library(here)
library(readxl)
library(lubridate)
library(grid)
library(gtable)
library(rlang)
library(ggplot2)
library(RplotterPkg)

current_dir <- here()
```

# Data
## Read in the Excel data
<div class="note">Note: The data was downloaded from the reference in the article.</div>

```{r}
data_file <- file.path(current_dir, "data/RedGreenGreyDots.xlsx")
if(!base::file.exists(data_file)){
  stop("Excel file RedGreenGreyDots.xlsx does not exist.")
}
data_dt <- readxl::read_xlsx(data_file)
data.table::setDT(data_dt)
str(data_dt)
```

<div class="note">Note: `data.table::setDT()` does an inplace conversion to data.table.</div>

# Transformations of data.table
## Add new columns with expressions in `j`
Add a new column *Movement15* that rounds times to nearest 15 minutes:
```{r}
data_dt[, Movement15 := lubridate::floor_date(MovementDateTime, "15 minutes")]
str(data_dt)
```


Add a new column *Counter* that uses the `ifelse` expression:
```{r}
data_dt[, Counter := ifelse(IN_OUT %chin% "IN", 1L, -1L)]
str(data_dt)
```

## Add a new column in `j` with an expression and grouping using `by`
Add a new column *SeqNo* using the `cumsum()` expression with grouping:
```{r}
data_dt[, SeqNo := cumsum(Counter), by = .(IN_OUT, Movement_Type,Staging_Post,Movement15)]
str(data_dt)
```

## Select rows in `i` and redefine a column value
For *Movement_Type* change all rows that contain "Transfer" to the value "Transfer":
```{r}
data_dt[Movement_Type %like% "Transfer", Movement_Type := "Transfer"]
str(data_dt)
```

# Plot the data
Plot the data using `RplotterPkg::multi_scatter_plot()` function:
```{r, fig.width=12, fig.height=10}
a_plot_table <- RplotterPkg::multi_scatter_plot(
  df = data_dt,
  factor_var = "Staging_Post",
  factor_x = "Movement15",
  aes_y = "SeqNo",
  aes_color = "Movement_Type",
  x_major_date_breaks = "3 hours",
  x_date_labels = "%H:%M",
  col_width = 6,
  display_plot = FALSE
)
grid::grid.newpage()
grid::grid.draw(a_plot_table)
```
