---
title: "Cheat Sheet for data.table"
output: 
   html_document:
    toc: yes
    toc_depth: 3
params:
  date: !r Sys.Date()    
---

```{r,setup, include=FALSE, eval=TRUE}
options(knitr.table.format = "html", width = 140)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 12, fig.height = 8)
```

<div>Author: Rick Dean</div>
<div>Article date: `r params$date`</div>

<div class="abstract">
  <p class="abstract">Abstract</p>
  The following contains notes/R scripts on learning data.table from a this [cheat sheet](https://s3.amazonaws.com/assets.datacamp.com/img/blog/data+table+cheat+sheet.pdf). Described as the official cheat sheet for the DataCamp course.
</div>  

```{r, warning=FALSE, message=FALSE}
library(knitr)
library(data.table)
```

# Data Analysis the Data.Table Way

## Example data.table

### Create the data.table *DT*
```{r}
DT <- data.table(
  V1 = c(1L, 2L),
  V2 = LETTERS[1:3],
  V3 = round(rnorm(4), 4),
  V4 = 1:12
)
DT
```
<div class="note">Note: the recycling</div>

### Define a function for resetting *DT*
```{r}
reset_DT = function(){
  DT <- data.table(
    V1 = c(1L, 2L),
    V2 = LETTERS[1:3],
    V3 = round(rnorm(4), 4),
    V4 = 1:12
  )
  return(DT)
}
```



## Subsetting rows using `i`
### By numbers
```{r}
DT[3:5,] # or DT[3:5]
```
### By column name
```{r}
DT[V2 == "A"] # selects all rows that have value "A"
```
### By multiple column name
```{r}
DT[V2 %in% c("A", "C")] # all rows that have "A" or "C" in column V2
```
## Manipulating on columns in `j`
### Select one column in `j`
```{r}
DT[, V2]
```
### Select several columns in `j` use `.()`
```{r}
DT[, .(V2, V3)] # returns a data.table
```
<div class="note">Note: `.()` is alias to `list()`. If `.()` is used the returned value is a *data.table*. If `.()` is not used the result is a vector.</div>
### Select columns without using `.()`
```{r}
DT[, c(V2,V3)] # returns a vector
```
### Call functions in `j`
```{r}
DT[, sum(V1)] # returns the sum of all elements in V1 as a vector
```
### Computing on several columns
```{r}
DT[, .(sum(V1), sd(V3))] # returns sum of V1 and sd of V3 in a data.table
```
### Assigning column names to computed columns
```{r}
DT[, .(Aggregate = sum(V1), Sd.V3 = sd(V3))]
```
### Columns get recycled if different lengths
```{r}
DT[, .(V1, Sd.V3 = sd(V3))] # Sd.V3 gets recycled
```
### Multiple expressions wrapped in curly braces
```{r}
DT[, {print(V2)  # like in-line funciton
      plot(V3)
      NULL}]
```
## Grouping with `j`
### Doing grouping with one column
```{r}
DT[, .(V4.Sum = sum(V4)), by = V1] # the sum of V4 for every group in V1
```
### Doing grouping with a function in `j`
```{r}
DT[, .(V4.mean = mean(V4)), by = V2] # the mean of V4 in subgroups of V2
```


### Doing grouping with several columns using `.()`
```{r}
DT[, .(V4.Sum = sum(V4)), by = .(V1,V2)]
```
### Call functions in `by`
```{r}
DT[, .(V4.Sum = sum(V4)), by = sign(V1-1)]
```
### Assigning new column names in `by`
```{r}
DT[, .(V4.Sum = sum(V4)), by = .(V1.01 = sign(V1-1))]
```
### Grouping on only a subset of rows by specifying `i`
```{r}
DT[1:5, .(V4.Sum = sum(V4)), by = V1]
```
### Using `.N` to get the total number of rows for each group
```{r}
DT[, .N, by = V1]
```
### Using `.N` in `j` returns the number of rows
```{r}
DT[,.N]
```


## Adding/Updating columns by reference in `j` using `:=`
### Updating one column using `:=`
```{r}
DT[, V1 := round(exp(V1), 2)] # V1 is updated
DT
```
### Adding/updating several columns (form 1)
```{r}
DT <- reset_DT()
DT[, c("V1", "V2") := .(round(exp(V1),2), rep(LETTERS[4:6],4))]
DT
```
### Adding/updating several columns (form 2 -- functional form)
```{r}
DT <- reset_DT()
DT[, `:=` (V1 = round(exp(V1),2), V2 = rep(LETTERS[4:6],4))]
DT
```
### Remove a column with `:=`
```{r}
DT[, V1 := NULL]
DT
```
### Remove several columns with `:=`
```{r}
DT <- reset_DT()
DT[, c("V1", "V2") := NULL]
DT
```
### Specify columns in a variable object
```{r}
DT <- reset_DT()
Cols.chosen <- c("V1", "V2")
DT[, (Cols.chosen) := NULL]
DT
```
<div class="note">Note: the use of `( )`</div>

## Indexing and keys
### Use `setkey()` to set key and sort specified column
```{r}
DT <- reset_DT()
setkey(DT, V2)
DT
```
### Use `key()` to identify the current key
```{r}
key(DT)
```

### Use keys to select rows for 1 value
```{r}
DT["A"]
```
### Use keys to select rows for several values
```{r}
DT[c("A", "C")]
```
### Use `mult` argument to control which row is returned
```{r}
DT["A", mult = "first"] # returns the first row of all rows that match the value "A" for V2
```
<div class="note">Note: there is also a "last" value for `mult`</div>

### Using the `nomatch` argument when selecting rows
```{r}
DT[c("A", "D"), nomatch = 0] # returns all rows with "A" or "D". Since no "D", it is not returned because of nomatch
```
### Result when `nomatch` is not specifed
```{r}
DT[c("A","D")]
```
### Multiple values for the key with a function in `j`
```{r}
DT[c("A","C"), sum(V4)] # returns the sum of V4 for rows of V2 with values "A", and "C"
```
### Using `.EACHI` in `by`
```{r}
DT[c("A","C"), sum(V4), by = .EACHI] # returns data.table with separate sums for V4 for rows with "A" or "C" under V2
```
### Multiple keys specified
```{r}
setkey(DT, V1, V2)
key(DT)
```
Now do row selection with these pair of keys:
```{r}
DT[.(2, "C")] # select rows with V1 = 2 and V2 = "C"
```
### Multiple keys and multiple values
```{r}
DT[.(2, c("A", "C"))]
```
## Advanced data table operations
### Using `.SD`
`.SD` is a data.table and holds all the values of all columns, except the one specified in `by`.

`.SD` is only accessible in `j`.
```{r}
DT[, print(.SD), by = V3] # to show what is in .SD
```
### `.SD` as a data.table of a subgroup
Get the first and last rows of each subgroup
```{r}
DT[, .SD[c(1, .N)], by = V2]
```
### Map across `.SD` with a function
```{r}
DT[, lapply(.SD, sum), by = V2] # sum of the rows of each subgroup of V2 for all of the columns
```

### Map across specific columns using `.SD` with a function
```{r}
DT[, lapply(.SD, sum), by = V2, .SDcols = c("V3", "V4")]
```

## Chaining statements
### With no chaining
1. Create intermediate data.table:
```{r}
DT_v4sum <-  DT[, .(V4.Sum = sum(V4)), by = V2]
DT_v4sum
```
2. Filter groups that have sums < 30:
```{r}
DT_v4sum <- DT_v4sum[V4.Sum < 30]
DT_v4sum
```
2. Order the result in descending order
```{r}
DT_v4sum <- DT_v4sum[,][order(-V4.Sum)]
DT_v4sum
```

### With chaining
```{r}
DT_v4sum <- DT[, .(V4.Sum = sum(V4)), by = V2][V4.Sum < 30][order(-V4.Sum)]
DT_v4sum
```
## Using the `set()` family
### `set()` to loop and update rows and columns
Syntax of `set()`:   for(i in from:to) set(DT, row, column, new value)
Example:
```{r}
rows <- list(3:4, 5:6)
cols <- 1:2
for(i in seq_along(rows)){
  set(DT, i = rows[[i]], j = cols[i], value = NA)
}
DT
```
### `setnames()` create or update column names
Syntax of `setnames(): setnames(DT, "old", "new")
```{r}
setnames(DT, c("V2", "V3"), c("V2.rating", "V3.datacamp"))
DT
```
### `setcolorder()` to reorder columns
Syntax of `setcolorder()`: setcolorder(DT, "neworder")
```{r}
DT <- reset_DT()
setcolorder(DT, c("V2","V1","V4","V3"))
DT
```

