---
title: "How data.table's fread can save you a lot of time and memory"
output: 
   html_document:
    toc: yes
    toc_depth: 3
    css: ../../style.css
params:
  date: !r Sys.Date()    
---

```{r,setup, include=FALSE, eval=TRUE}
options(knitr.table.format = "html")
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 12, fig.height = 8)
```

<div>Author: Rick Dean</div>
<div>Article date: `r params$date`</div>

<div class="abstract">
  <p class="abstract">Abstract</p>
The following notes/R scripts are inspired on an article by Jozef Hajnala,
[How data.table’s fread can save you a lot of time and memory, and take input from shell commands"](https://jozef.io/r917-fread-comparisons/)
</div>  

```{r, warning=FALSE, message=FALSE}
library(knitr)
library(readr)
library(data.table)
library(here)
library(tictoc)
library(pryr)

current_dir <- here()
```

# Compare Timing `utils::read.csv()`, `readr::read_csv()`, and `data.table::fread()`
## Establish data directory

<div class="note">Note: We are using the [Airline on-time performance data](http://stat-computing.org/dataexpo/2009/the-data.html) referenced in the article.</div>

Establish source directory for the source airline data file:
```{r}
source_file <- file.path(current_dir, "data/january_2020.csv")
if(!base::file.exists(source_file)){
  stop("Source airline file does not exist.")
}
```

## Using `utils::read.csv()`
Read and time airline data using `utils::read.csv()`
```{r}
tictoc::tic("utils::read.csv()")
base_airline_df <- utils::read.csv(source_file)
tictoc::toc()
```
## Using `readr::read_csv()`
Read and time airline data using `readr::read_csv()`
```{r}
tictoc::tic("readr::read_csv()")
readr_airline_df <- readr::read_csv(source_file, progress = F)
tictoc::toc()
```
## Using `data.table::fread()`
Read and time airline data using `data.table::fread()`
```{r}
tictoc::tic("data.table::fread()")
fread_airline_df <- data.table::fread(source_file)
tictoc::toc()
```

# Compare memory usage
Memory change for `utils::read.csv()`
```{r}
pryr::mem_change(utils::read.csv(source_file))
```

Memory change for `readr::read_csv()`
```{r}
pryr::mem_change(readr::read_csv(source_file))
```

Memory change for `data.table::fread()`
```{r}
pryr::mem_change(data.table::fread(source_file))
```

# Compare data frame size
Data frame size for `utils::read.csv()`
```{r}
pryr::object_size(base_airline_df)
```

Data frame size for `readr::read_csv()`
```{r}
pryr::object_size(readr_airline_df)
```

Data frame size for `data.table::fread()`
```{r}
pryr::object_size(fread_airline_df)
```
